---
title: "MiCBuS — Ovarian HGSOC notebook (source)"
subtitle: "Reduced V2 local R Markdown notebook, ported to Quarto. Source only — figures live in the [rendered report](report.html)."
date: 2025-05-09
toc: true
number-sections: true
format:
  html:
    toc-location: right
    code-fold: false
    code-tools: true
execute:
  enabled: false
params:
  mode: "v2_local"
  install_missing: true
  holdout_class: "stromal_caf"
  optional_holdout_class: "myeloid"
  max_cells_per_class_smoke: 400
  max_cells_per_class_reduced: 1000
  max_cells_per_class_v2: 2000
  tcga_samples_smoke: 20
  tcga_samples_reduced: 75
  tcga_samples_v2: 150
  pseudo_bulk_reps_smoke: 35
  pseudo_bulk_reps_reduced: 150
  pseudo_bulk_reps_v2: 300
  visium_samples: ["SP1", "SP2"]
  seed: 42
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  fig.width = 9,
  fig.height = 6
)

set.seed(params$seed)

required_packages <- c(
  "data.table", "dplyr", "ggplot2", "Matrix", "patchwork",
  "pheatmap", "Seurat", "tibble", "tidyr"
)

optional_packages <- c("GEOquery", "TCGAbiolinks", "SummarizedExperiment")
bioc_packages <- c("GEOquery", "TCGAbiolinks", "SummarizedExperiment")
cran_packages <- setdiff(c(required_packages, optional_packages, "BiocManager", "remotes"), bioc_packages)

if (isTRUE(params$install_missing)) {
  missing_cran <- cran_packages[!vapply(cran_packages, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing_cran) > 0) {
    install.packages(missing_cran, repos = "https://cloud.r-project.org")
  }

  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager", repos = "https://cloud.r-project.org")
  }

  missing_bioc <- bioc_packages[!vapply(bioc_packages, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing_bioc) > 0) {
    BiocManager::install(missing_bioc, ask = FALSE, update = FALSE)
  }

  if (!requireNamespace("MiCBuS", quietly = TRUE) && requireNamespace("remotes", quietly = TRUE)) {
    try(
      remotes::install_github("Shanshan-Zhang/MiCBuS", upgrade = "never"),
      silent = TRUE
    )
  }
}

missing_required <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_required) > 0) {
  stop(
    "Install required packages before rendering: ",
    paste(missing_required, collapse = ", ")
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))

dir.create("data/raw/GSE211956", recursive = TRUE, showWarnings = FALSE)
dir.create("data/raw/TCGA_OV", recursive = TRUE, showWarnings = FALSE)
dir.create("data/cache", recursive = TRUE, showWarnings = FALSE)
dir.create("results", recursive = TRUE, showWarnings = FALSE)

is_smoke <- identical(params$mode, "smoke")
is_v2_local <- identical(params$mode, "v2_local")
max_cells_per_class <- if (is_v2_local) {
  params$max_cells_per_class_v2
} else if (is_smoke) {
  params$max_cells_per_class_smoke
} else {
  params$max_cells_per_class_reduced
}
n_tcga_samples <- if (is_v2_local) {
  params$tcga_samples_v2
} else if (is_smoke) {
  params$tcga_samples_smoke
} else {
  params$tcga_samples_reduced
}
n_pseudobulk <- if (is_v2_local) {
  params$pseudo_bulk_reps_v2
} else if (is_smoke) {
  params$pseudo_bulk_reps_smoke
} else {
  params$pseudo_bulk_reps_reduced
}
```

# Scope

This notebook is a report-grade V2 local ovarian HGSOC run for testing whether
MiCBuS can recover marker genes for a tumor microenvironment population that is
present in bulk tumors but intentionally omitted from the single-cell reference.

The default missing population is `stromal_caf`. The expected positive controls
are `COL1A1`, `COL1A2`, `DCN`, `LUM`, `FAP`, `ACTA2`, `PDGFRA`, and `PDGFRB`.
An optional second holdout is `myeloid`, with expected markers `LYZ`, `C1QA`,
`C1QB`, `C1QC`, `CD68`, `APOE`, and `SPP1`.

This workflow uses only public processed expression data:

- `GSE211956` processed GEO files for five scRNA-seq samples and two
  Visium sections.
- Open TCGA-OV summarized gene expression.
- No raw reads, dbGaP, EGA, low-pass WGS, CNA calls, or restricted data.

# Helper functions

```{r helpers}
`%||%` <- function(x, y) {
  if (is.null(x) || length(x) == 0) y else x
}

safe_genes <- function(genes, object_or_matrix) {
  available <- if (inherits(object_or_matrix, "Seurat")) {
    rownames(object_or_matrix)
  } else {
    rownames(object_or_matrix)
  }
  intersect(genes, available)
}

join_rna_layers <- function(seurat_object) {
  join_fn <- NULL
  if ("JoinLayers" %in% getNamespaceExports("Seurat")) {
    join_fn <- getExportedValue("Seurat", "JoinLayers")
  } else if (requireNamespace("SeuratObject", quietly = TRUE) &&
             "JoinLayers" %in% getNamespaceExports("SeuratObject")) {
    join_fn <- getExportedValue("SeuratObject", "JoinLayers")
  }

  if (!is.null(join_fn)) {
    seurat_object <- join_fn(seurat_object, assay = "RNA")
  }
  seurat_object
}

get_rna_data <- function(seurat_object) {
  tryCatch(
    Seurat::GetAssayData(seurat_object, assay = "RNA", layer = "data"),
    error = function(e) {
      Seurat::GetAssayData(seurat_object, assay = "RNA", slot = "data")
    }
  )
}

marker_sets <- list(
  malignant_epithelial = c("EPCAM", "KRT8", "KRT18", "KRT19", "PAX8", "MUC16", "MSLN"),
  stromal_caf = c("COL1A1", "COL1A2", "COL3A1", "DCN", "LUM", "FAP", "ACTA2", "PDGFRA", "PDGFRB"),
  myeloid = c("PTPRC", "LYZ", "C1QA", "C1QB", "C1QC", "CD68", "APOE", "SPP1"),
  t_nk = c("PTPRC", "CD3D", "CD3E", "TRAC", "NKG7", "GNLY", "KLRD1"),
  b_plasma = c("PTPRC", "MS4A1", "CD79A", "CD79B", "MZB1", "JCHAIN", "IGHG1"),
  endothelial = c("PECAM1", "VWF", "KDR", "CDH5", "ENG", "PLVAP")
)

positive_controls <- list(
  stromal_caf = c("COL1A1", "COL1A2", "DCN", "LUM", "FAP", "ACTA2", "PDGFRA", "PDGFRB"),
  myeloid = c("LYZ", "C1QA", "C1QB", "C1QC", "CD68", "APOE", "SPP1")
)

read_count_txt <- function(path) {
  dt <- data.table::fread(path)
  gene_col <- names(dt)[1]
  genes <- make.unique(as.character(dt[[gene_col]]))
  mat <- as.matrix(dt[, -1, with = FALSE])
  rownames(mat) <- genes
  Matrix::Matrix(mat, sparse = TRUE)
}

sample_id_from_path <- function(path) {
  base <- basename(path)
  sub("^GSM[0-9]+_counts_([^.]*)\\.txt\\.gz$", "\\1", base)
}

downsample_by_class <- function(meta, class_col, max_per_class, seed = 1) {
  set.seed(seed)
  split(seq_len(nrow(meta)), meta[[class_col]]) |>
    lapply(function(idx) sample(idx, min(length(idx), max_per_class))) |>
    unlist(use.names = FALSE)
}

average_expression_by_class <- function(seurat_object, class_col = "broad_class") {
  avg <- Seurat::AverageExpression(
    seurat_object,
    group.by = class_col,
    assays = "RNA",
    slot = "data"
  )$RNA
  as.matrix(avg)
}

score_gene_set_matrix <- function(expr, genes) {
  genes <- intersect(genes, rownames(expr))
  if (length(genes) == 0) {
    stop("No marker genes were found in the expression matrix.")
  }
  colMeans(expr[genes, , drop = FALSE])
}

normalize_bulk_matrix <- function(counts_or_expression) {
  mat <- as.matrix(counts_or_expression)
  mat <- mat[rowSums(is.finite(mat)) == ncol(mat), , drop = FALSE]
  mat <- mat[!duplicated(rownames(mat)), , drop = FALSE]
  if (max(mat, na.rm = TRUE) > 100) {
    lib <- colSums(mat)
    mat <- t(t(mat) / lib * 1e6)
    mat <- log2(mat + 1)
  }
  mat
}

query_tcga_ov_counts <- function(n_samples, seed = 1) {
  workflows <- c("STAR - Counts", "HTSeq - Counts")
  last_error <- NULL

  for (workflow in workflows) {
    query <- tryCatch(
      TCGAbiolinks::GDCquery(
        project = "TCGA-OV",
        data.category = "Transcriptome Profiling",
        data.type = "Gene Expression Quantification",
        workflow.type = workflow,
        sample.type = "Primary Tumor"
      ),
      error = function(e) e
    )

    if (inherits(query, "error")) {
      last_error <- query
      next
    }

    cases_result <- TCGAbiolinks::getResults(query, cols = "cases")
    cases <- if (is.data.frame(cases_result)) cases_result$cases else cases_result
    cases <- unique(cases)
    cases <- cases[!is.na(cases)]
    if (length(cases) == 0) {
      next
    }

    set.seed(seed)
    selected_cases <- sample(cases, min(n_samples, length(cases)))

    subset_query <- TCGAbiolinks::GDCquery(
      project = "TCGA-OV",
      data.category = "Transcriptome Profiling",
      data.type = "Gene Expression Quantification",
      workflow.type = workflow,
      sample.type = "Primary Tumor",
      barcode = selected_cases
    )

    return(subset_query)
  }

  stop("Unable to create a TCGA-OV query. Last error: ", last_error$message %||% "none")
}

simulate_dirichlet_weights <- function(n, k, alpha = 1) {
  raw <- matrix(rgamma(n * k, shape = alpha, rate = 1), nrow = n, ncol = k)
  raw / rowSums(raw)
}

run_micbus_fallback <- function(bulk_expr, reference_profiles, n_reps, seed = 1) {
  set.seed(seed)
  genes <- Reduce(intersect, list(rownames(bulk_expr), rownames(reference_profiles)))
  bulk <- bulk_expr[genes, , drop = FALSE]
  ref <- reference_profiles[genes, , drop = FALSE]

  weights <- simulate_dirichlet_weights(n_reps, ncol(ref), alpha = 1)
  pseudo <- ref %*% t(weights)
  colnames(pseudo) <- paste0("pseudo_", seq_len(n_reps))

  combined <- cbind(bulk, pseudo)
  group <- factor(c(rep("bulk", ncol(bulk)), rep("pseudo", ncol(pseudo))), levels = c("pseudo", "bulk"))

  res <- lapply(seq_len(nrow(combined)), function(i) {
    fit <- tryCatch(
      stats::t.test(combined[i, group == "bulk"], combined[i, group == "pseudo"]),
      error = function(e) NULL
    )
    if (is.null(fit)) {
      return(data.frame(gene = rownames(combined)[i], logFC = NA_real_, p_value = NA_real_))
    }
    data.frame(
      gene = rownames(combined)[i],
      logFC = mean(combined[i, group == "bulk"]) - mean(combined[i, group == "pseudo"]),
      p_value = fit$p.value
    )
  }) |>
    dplyr::bind_rows() |>
    dplyr::mutate(p_adj = p.adjust(p_value, method = "BH")) |>
    dplyr::arrange(dplyr::desc(logFC), p_adj)

  res
}

run_micbus <- function(bulk_expr, incomplete_reference, n_reps, seed = 1) {
  # Prefer a user-installed MiCBuS package when it exposes a recognizable entry point.
  if (requireNamespace("MiCBuS", quietly = TRUE)) {
    ns <- asNamespace("MiCBuS")
    exported <- getNamespaceExports("MiCBuS")
    candidate <- intersect(c("MiCBuS", "micbus", "run_micbus", "runMiCBuS"), exported)

    if (length(candidate) > 0) {
      fn <- get(candidate[1], envir = ns)
      message("Running MiCBuS package function: MiCBuS::", candidate[1])
      out <- tryCatch(
        fn(
          bulk = bulk_expr,
          sc_ref = incomplete_reference,
          n_pseudobulk = n_reps,
          seed = seed
        ),
        error = function(e) e
      )
      if (!inherits(out, "error")) {
        return(as.data.frame(out))
      }
      message("Installed MiCBuS call failed; using tutorial fallback. Error: ", out$message)
    }
  }

  message("Using tutorial fallback: Dirichlet pseudo-bulk plus bulk-vs-pseudo differential test.")
  run_micbus_fallback(bulk_expr, incomplete_reference, n_reps = n_reps, seed = seed)
}
```

# Download public processed data

`GSE211956` contains five processed 10x scRNA-seq count matrices and eight
processed Space Ranger Visium samples. The code below downloads the GEO
supplementary archive through `GEOquery` and extracts only processed files.

```{r download-geo}
geo_dir <- "data/raw/GSE211956"
raw_tar <- file.path(geo_dir, "GSE211956_RAW.tar")

if (!file.exists(raw_tar)) {
  if (!requireNamespace("GEOquery", quietly = TRUE)) {
    stop("Install GEOquery to download GSE211956 processed files.")
  }
  GEOquery::getGEOSuppFiles("GSE211956", baseDir = geo_dir, makeDirectory = FALSE)
}

if (file.exists(raw_tar) && length(list.files(geo_dir, pattern = "GSM.*", full.names = TRUE)) == 0) {
  utils::untar(raw_tar, exdir = geo_dir)
}

scrna_count_files <- list.files(
  geo_dir,
  pattern = "GSM[0-9]+_counts_.*\\.txt\\.gz$",
  full.names = TRUE
)

visium_prefixes <- unlist(params$visium_samples)
visium_files <- list.files(geo_dir, pattern = "GSM.*_SP[0-9]+_.*\\.(gz|zip)$", full.names = TRUE)

length(scrna_count_files)
basename(scrna_count_files)
```

# Build the V2 local scRNA reference

The GEO scRNA files are processed count matrices without harmonized broad cell
class labels. This report assigns broad labels by clustering the merged object
and selecting the strongest canonical marker module for each cluster.

```{r scrna-reference}
scrna_cache <- file.path(
  "data/cache",
  paste0("gse211956_scrna_", params$mode, "_", max_cells_per_class, ".rds")
)

if (file.exists(scrna_cache)) {
  scrna <- readRDS(scrna_cache)
} else {
  if (length(scrna_count_files) == 0) {
    stop("No GSE211956 scRNA count files were found after GEO download.")
  }

  objects <- lapply(scrna_count_files, function(path) {
    sample_id <- sample_id_from_path(path)
    counts <- read_count_txt(path)
    obj <- Seurat::CreateSeuratObject(counts, project = sample_id, min.cells = 3, min.features = 200)
    obj$sample_id <- sample_id
    obj[["percent.mt"]] <- Seurat::PercentageFeatureSet(obj, pattern = "^MT-")
    subset(obj, subset = nFeature_RNA >= 500 & nCount_RNA >= 1000 & percent.mt < 20)
  })

  scrna <- Reduce(function(x, y) merge(x, y), objects)
  scrna <- Seurat::NormalizeData(scrna)
  scrna <- Seurat::FindVariableFeatures(scrna, nfeatures = 2500)
  scrna <- Seurat::ScaleData(scrna, features = Seurat::VariableFeatures(scrna))
  scrna <- Seurat::RunPCA(scrna, features = Seurat::VariableFeatures(scrna), npcs = 30)
  scrna <- Seurat::FindNeighbors(scrna, dims = 1:20)
  scrna <- Seurat::FindClusters(scrna, resolution = 0.5)
  scrna <- Seurat::RunUMAP(scrna, dims = 1:20)

  for (nm in names(marker_sets)) {
    genes <- safe_genes(marker_sets[[nm]], scrna)
    if (length(genes) > 0) {
      scrna <- Seurat::AddModuleScore(scrna, features = list(genes), name = paste0(nm, "_score"))
    }
  }

  score_cols <- paste0(names(marker_sets), "_score1")
  cluster_scores <- scrna@meta.data |>
    tibble::rownames_to_column("cell") |>
    dplyr::group_by(seurat_clusters) |>
    dplyr::summarise(dplyr::across(dplyr::all_of(score_cols), mean), .groups = "drop")

  cluster_label <- apply(cluster_scores[, score_cols, drop = FALSE], 1, function(x) {
    names(marker_sets)[which.max(x)]
  })
  names(cluster_label) <- cluster_scores$seurat_clusters

  scrna$broad_class <- unname(cluster_label[as.character(scrna$seurat_clusters)])

  keep_idx <- downsample_by_class(
    scrna@meta.data,
    class_col = "broad_class",
    max_per_class = max_cells_per_class,
    seed = params$seed
  )
  scrna <- subset(scrna, cells = colnames(scrna)[keep_idx])
  scrna <- join_rna_layers(scrna)

  saveRDS(scrna, scrna_cache)
}

table(scrna$broad_class)
```

```{r broad-class-umap}
Seurat::DimPlot(scrna, group.by = "broad_class", label = TRUE, repel = TRUE) +
  ggplot2::ggtitle("Reduced GSE211956 scRNA reference")
```

# Complete and incomplete references

MiCBuS is tested by deliberately removing the holdout class from the scRNA
reference. The complete object remains available only for validation.

```{r complete-incomplete-reference}
if (!exists("join_rna_layers")) {
  join_rna_layers <- function(seurat_object) {
    join_fn <- NULL
    if ("JoinLayers" %in% getNamespaceExports("Seurat")) {
      join_fn <- getExportedValue("Seurat", "JoinLayers")
    } else if (requireNamespace("SeuratObject", quietly = TRUE) &&
               "JoinLayers" %in% getNamespaceExports("SeuratObject")) {
      join_fn <- getExportedValue("SeuratObject", "JoinLayers")
    }

    if (!is.null(join_fn)) {
      seurat_object <- join_fn(seurat_object, assay = "RNA")
    }
    seurat_object
  }
}

holdout_class <- params$holdout_class
if (!holdout_class %in% scrna$broad_class) {
  stop("Holdout class not found in broad labels: ", holdout_class)
}

complete_ref <- scrna
incomplete_ref <- subset(scrna, subset = broad_class != holdout_class)
complete_ref <- join_rna_layers(complete_ref)
incomplete_ref <- join_rna_layers(incomplete_ref)

complete_profiles <- average_expression_by_class(complete_ref)
incomplete_profiles <- average_expression_by_class(incomplete_ref)

colnames(complete_profiles)
colnames(incomplete_profiles)
```

# TCGA-OV bulk expression

The bulk expression input is open TCGA-OV primary tumor expression. The notebook
uses `TCGAbiolinks` to obtain HTSeq counts from GDC and then keeps a small,
deterministic subset for smoke testing or reduced local rendering.

```{r tcga-download}
tcga_cache <- file.path("data/cache", paste0("tcga_ov_bulk_", n_tcga_samples, ".rds"))

if (file.exists(tcga_cache)) {
  bulk_expr <- readRDS(tcga_cache)
} else {
  if (!requireNamespace("TCGAbiolinks", quietly = TRUE) ||
      !requireNamespace("SummarizedExperiment", quietly = TRUE)) {
    stop("Install TCGAbiolinks and SummarizedExperiment to download TCGA-OV bulk expression.")
  }

  query <- query_tcga_ov_counts(n_tcga_samples, seed = params$seed)
  TCGAbiolinks::GDCdownload(query, directory = "data/raw/TCGA_OV")
  se <- TCGAbiolinks::GDCprepare(query, directory = "data/raw/TCGA_OV")

  counts <- SummarizedExperiment::assay(se)
  rd <- as.data.frame(SummarizedExperiment::rowData(se))
  gene_symbols <- rd$external_gene_name %||% rd$gene_name %||% rd$gene_id
  keep <- !is.na(gene_symbols) & gene_symbols != ""
  counts <- counts[keep, , drop = FALSE]
  gene_symbols <- gene_symbols[keep]

  counts_by_symbol <- rowsum(as.matrix(counts), group = gene_symbols)
  bulk_expr <- normalize_bulk_matrix(counts_by_symbol)

  saveRDS(bulk_expr, tcga_cache)
}

dim(bulk_expr)
```

# Run MiCBuS

The primary call uses an installed `MiCBuS` package if it is present and exposes
a recognizable exported function. If not, the tutorial falls back to the core
teaching idea: generate Dirichlet pseudo-bulk mixtures from the incomplete
reference and identify genes higher in TCGA bulk tumors than in those mixtures.

```{r run-micbus}
common_genes <- Reduce(intersect, list(rownames(bulk_expr), rownames(incomplete_profiles)))
bulk_for_micbus <- bulk_expr[common_genes, , drop = FALSE]
ref_for_micbus <- incomplete_profiles[common_genes, , drop = FALSE]

micbus_res <- run_micbus(
  bulk_expr = bulk_for_micbus,
  incomplete_reference = ref_for_micbus,
  n_reps = n_pseudobulk,
  seed = params$seed
)

if (!"gene" %in% names(micbus_res)) {
  micbus_res <- tibble::rownames_to_column(as.data.frame(micbus_res), "gene")
}

score_col <- intersect(c("score", "statistic", "logFC", "avg_log2FC"), names(micbus_res))[1]
padj_col <- intersect(c("p_adj", "padj", "adj.P.Val", "q_value", "FDR"), names(micbus_res))[1]

if (is.na(score_col)) {
  numeric_cols <- names(micbus_res)[vapply(micbus_res, is.numeric, logical(1))]
  score_col <- numeric_cols[1]
}

top_psmarkers <- micbus_res |>
  dplyr::filter(!is.na(.data[[score_col]])) |>
  dplyr::arrange(dplyr::desc(.data[[score_col]])) |>
  dplyr::slice_head(n = 50)

utils::write.csv(micbus_res, "results/v2_ovarian_micbus_psmarkers.csv", row.names = FALSE)

top_psmarkers |>
  dplyr::select(dplyr::any_of(c("gene", score_col, padj_col))) |>
  dplyr::slice_head(n = 20)
```

# Held-out scRNA validation

If MiCBuS is behaving as expected, top pseudo-marker genes should be enriched in
the omitted broad class when the complete scRNA object is used for validation.

```{r heldout-de}
if (!exists("join_rna_layers")) {
  join_rna_layers <- function(seurat_object) {
    join_fn <- NULL
    if ("JoinLayers" %in% getNamespaceExports("Seurat")) {
      join_fn <- getExportedValue("Seurat", "JoinLayers")
    } else if (requireNamespace("SeuratObject", quietly = TRUE) &&
               "JoinLayers" %in% getNamespaceExports("SeuratObject")) {
      join_fn <- getExportedValue("SeuratObject", "JoinLayers")
    }

    if (!is.null(join_fn)) {
      seurat_object <- join_fn(seurat_object, assay = "RNA")
    }
    seurat_object
  }
}

Idents(complete_ref) <- complete_ref$broad_class
complete_ref <- join_rna_layers(complete_ref)
heldout_markers <- Seurat::FindMarkers(
  complete_ref,
  ident.1 = holdout_class,
  ident.2 = setdiff(unique(complete_ref$broad_class), holdout_class),
  only.pos = TRUE,
  min.pct = 0.1,
  logfc.threshold = 0.1
) |>
  tibble::rownames_to_column("gene") |>
  dplyr::arrange(dplyr::desc(avg_log2FC))

top_micbus_genes <- top_psmarkers$gene
overlap <- intersect(top_micbus_genes, heldout_markers$gene[seq_len(min(100, nrow(heldout_markers)))])
expected <- positive_controls[[holdout_class]] %||% character(0)

validation_summary <- tibble::tibble(
  metric = c("top_50_micbus_genes", "overlap_with_top_100_heldout_de", "expected_marker_recovery"),
  value = c(
    length(top_micbus_genes),
    length(overlap),
    length(intersect(top_micbus_genes, expected))
  )
)

validation_summary
intersect(top_micbus_genes, expected)
```

```{r heldout-dotplot}
plot_genes <- unique(c(
  intersect(top_micbus_genes, rownames(complete_ref))[seq_len(min(12, length(intersect(top_micbus_genes, rownames(complete_ref)))))],
  intersect(expected, rownames(complete_ref))
))

Seurat::DotPlot(complete_ref, features = plot_genes, group.by = "broad_class") +
  ggplot2::coord_flip() +
  ggplot2::ggtitle("Top MiCBuS genes and expected holdout markers across broad classes")
```

```{r heldout-heatmap}
avg_complete <- average_expression_by_class(complete_ref)
heat_genes <- intersect(plot_genes, rownames(avg_complete))
pheatmap::pheatmap(
  avg_complete[heat_genes, , drop = FALSE],
  scale = "row",
  cluster_cols = FALSE,
  main = "Complete-reference average expression"
)
```

# Bulk sanity check

The top pseudo-marker score should vary across TCGA-OV tumors. This is a sanity
check only; subtype validation is optional and intentionally not required here.

```{r bulk-score}
bulk_score_genes <- intersect(top_micbus_genes[seq_len(min(25, length(top_micbus_genes)))], rownames(bulk_expr))
bulk_scores <- tibble::tibble(
  sample = colnames(bulk_expr),
  psmarker_score = score_gene_set_matrix(bulk_expr, bulk_score_genes)
)

ggplot2::ggplot(bulk_scores, ggplot2::aes(x = reorder(sample, psmarker_score), y = psmarker_score)) +
  ggplot2::geom_col(fill = "#3d6f8e") +
  ggplot2::coord_flip() +
  ggplot2::labs(x = "TCGA-OV sample", y = "Mean top psMarker expression") +
  ggplot2::theme_minimal(base_size = 11)
```

# Spatial validation with two Visium sections

The Visium step scores spots for the top pseudo-marker panel and plots the score
in tissue coordinates. This is spatial expression validation only; it does not
make CNA or subclone claims.

```{r load-visium}
load_visium_sample <- function(prefix, geo_dir) {
  sample_files <- list.files(geo_dir, pattern = paste0("_", prefix, "_"), full.names = TRUE)
  matrix_file <- sample_files[grepl("_matrix\\.mtx\\.gz$", sample_files)][1]
  feature_file <- sample_files[grepl("_features\\.tsv\\.gz$", sample_files)][1]
  barcode_file <- sample_files[grepl("_barcodes\\.tsv\\.gz$", sample_files)][1]
  spatial_zip <- sample_files[grepl("_spatial\\.zip$", sample_files)][1]

  if (any(is.na(c(matrix_file, feature_file, barcode_file, spatial_zip)))) {
    stop("Missing Visium processed files for ", prefix)
  }

  spatial_dir <- file.path(geo_dir, paste0(prefix, "_spatial"))
  if (!dir.exists(spatial_dir)) {
    dir.create(spatial_dir, recursive = TRUE)
    utils::unzip(spatial_zip, exdir = spatial_dir)
  }

  mat <- Matrix::readMM(gzfile(matrix_file))
  features <- data.table::fread(feature_file, header = FALSE)
  barcodes <- data.table::fread(barcode_file, header = FALSE)
  rownames(mat) <- make.unique(features[[2]])
  colnames(mat) <- barcodes[[1]]

  obj <- Seurat::CreateSeuratObject(mat, project = prefix, min.features = 100)
  obj <- Seurat::NormalizeData(obj)

  tissue_positions <- list.files(spatial_dir, pattern = "tissue_positions.*\\.csv$", recursive = TRUE, full.names = TRUE)[1]
  if (!is.na(tissue_positions)) {
    pos <- data.table::fread(tissue_positions, header = FALSE)
    if (ncol(pos) >= 6) {
      colnames(pos)[1:6] <- c("barcode", "in_tissue", "array_row", "array_col", "pxl_row", "pxl_col")
      pos <- as.data.frame(pos)
      rownames(pos) <- pos$barcode
      obj <- AddMetaData(obj, pos[colnames(obj), c("in_tissue", "array_row", "array_col", "pxl_row", "pxl_col")])
    }
  }

  obj$visium_sample <- prefix
  obj
}

visium_objects <- lapply(visium_prefixes, load_visium_sample, geo_dir = geo_dir)
names(visium_objects) <- visium_prefixes
```

```{r spatial-score}
if (!exists("get_rna_data")) {
  get_rna_data <- function(seurat_object) {
    tryCatch(
      Seurat::GetAssayData(seurat_object, assay = "RNA", layer = "data"),
      error = function(e) {
        Seurat::GetAssayData(seurat_object, assay = "RNA", slot = "data")
      }
    )
  }
}

spatial_score_genes <- intersect(bulk_score_genes, rownames(visium_objects[[1]]))

spatial_plots <- lapply(names(visium_objects), function(prefix) {
  obj <- visium_objects[[prefix]]
  genes <- intersect(spatial_score_genes, rownames(obj))
  expr <- get_rna_data(obj)
  obj$psmarker_score <- score_gene_set_matrix(expr, genes)

  meta <- obj@meta.data |>
    tibble::rownames_to_column("barcode") |>
    dplyr::filter(!is.na(pxl_col), !is.na(pxl_row))

  ggplot2::ggplot(meta, ggplot2::aes(x = pxl_col, y = -pxl_row, color = psmarker_score)) +
    ggplot2::geom_point(size = 1.2) +
    ggplot2::coord_fixed() +
    ggplot2::scale_color_viridis_c(option = "magma") +
    ggplot2::labs(title = paste0(prefix, " top psMarker score"), x = NULL, y = NULL, color = "score") +
    ggplot2::theme_void()
})

Reduce(`+`, spatial_plots)
```

```{r spatial-feature-panel}
if (!exists("get_rna_data")) {
  get_rna_data <- function(seurat_object) {
    tryCatch(
      Seurat::GetAssayData(seurat_object, assay = "RNA", layer = "data"),
      error = function(e) {
        Seurat::GetAssayData(seurat_object, assay = "RNA", slot = "data")
      }
    )
  }
}

feature_genes <- intersect(c(expected, top_micbus_genes), rownames(visium_objects[[1]]))
feature_genes <- feature_genes[seq_len(min(6, length(feature_genes)))]

feature_plots <- lapply(feature_genes, function(gene) {
  obj <- visium_objects[[1]]
  expr <- get_rna_data(obj)
  meta <- obj@meta.data |>
    tibble::rownames_to_column("barcode") |>
    dplyr::filter(!is.na(pxl_col), !is.na(pxl_row)) |>
    dplyr::mutate(expression = as.numeric(expr[gene, barcode]))

  ggplot2::ggplot(meta, ggplot2::aes(x = pxl_col, y = -pxl_row, color = expression)) +
    ggplot2::geom_point(size = 1.0) +
    ggplot2::coord_fixed() +
    ggplot2::scale_color_viridis_c(option = "plasma") +
    ggplot2::labs(title = gene, x = NULL, y = NULL, color = "expr") +
    ggplot2::theme_void()
})

patchwork::wrap_plots(feature_plots, ncol = 3)
```

# Optional second holdout

This section does not rerun the full notebook by default. To test the myeloid
holdout, render with `params$holdout_class = "myeloid"` and check recovery of
`LYZ`, `C1QA`, `C1QB`, `C1QC`, `CD68`, `APOE`, and `SPP1`.

```{r session-info}
sessionInfo()
```
